Simulating High-Dimensional Multivariate Data

### 2020-12-17 030-bigsimr-package.Rmd
4th draft. Address WWP ch030 part 2

We'll demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.

@@

For simplicity and to provide a minimal working example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.

@@

Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
We aim to simulate random two-component vectors mimicking this structure. 
The margins are not normally distributed, particularly the ozone level exhibits a strong positive skew.

```{r ch030-aq-joint-dist, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Bivariate scatterplot of Ozone vs. Temperature with estimated marginal densities. The data are left skewed tails and appear to be correlation.", cache=F}

@@

Next, we specify the marginal distributions and correlation coefficient (both its type and magnitude).
Here the analyst is free to be creative.
For this example, we will not take up goodness-of-fit considerations to determine the marginal distributions.
But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.

*Specifying marginal distributions*.
Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we'll assume `Temp` is normally distributed and `Ozone` is log-normally distributed since its values are positive and skewed.
We'll use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

@@

Next, we'll configure the input marginals for later input into `bigsimr::rvec`.
The marginal distributions are specifying using `R`'s special `alist` function.
This allows one to enter the distributions without evaluating anything (yet).

@@

Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ enter into the `bigsimr::rvec` algorithm.
This implementation strategy supports all `R` `base` probability distributions.
And allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).

It is a bit inconvenient to have to fill in the parameter values manually each time, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it will evaluate the right hand side of argument values within the list.
This is intended to help when scaling up your code to high dimensions when many marginals to specify.

@@

*Specifying correlation*.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) for evaluation strategies and guidance).
To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
Notably, these estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see [Conclusion and Discussion]({#discussion) for more on this).

@@

*Checking the theoretical correlation bounds*
As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
To ensure that the simulation is not configured to impossible settings, we provide the `bigsimr::cor_bounds` function provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).

@@

*Simulating random vectors*.
Finally, we arrive at the main function of `bigsimr`, `rvec`.
Let's now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.

@@

*Creating a custom marginal distribution*.
Because `bigsimr` uses an `alist` to store the margins, any probability distribution with a well-defined inverse CDF can be used including custom marginal distributions not provided in base `R`. 
Therefore, to specify a marginal distribution absent from an existing package, the user needs to provide a closed-form expression to form the corresponding quantile function.

It is important to follow `R`'s naming convention for probability distributions.
Users should prefix distributions with `r` and `q` for *random* and *quantile* respectively.
Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, it is required to supply univariate random number generator as well.

For an example, let's provide a custom *Pareto* distribution for use with `bigsimr`.
The Pareto CDF is

@@

Now with the `qpareto` and `rpareto` functions, we can use the distribution in `bigsimr` just like the other built-in distributions.

@@

*Using `bigsimr` on a computing cluster via `rslurm`*.
Though `bigsimr` runs quickly, at large $d$ users may want to run jobs on a shared computing server.
The R package `rslurm` makes it easy to run embarrassingly large parallel `rvec` calls.
This example assumes that `bigsimr` is installed on a system with a slurm scheduler installed.
A single run of `bigsimr:rvec` using `rslurm` can be code as:

@@

Now, let's show off the real power of combining `bigsimr` and `rslurm` by simulating many correlation structures.
The `rslurm::slumr_map` syntax mirrors the `base::lapply` and `purrr::map` functions.

@@

On a cluster carrying 24 nodes with 48 threads, these 100 jobs completed in about a minute.

>>>>

We demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package.
First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.

@@

For simplicity and to provide a minimal working example, we consider bivariate simulation of the two `airquality` variables `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.

@@

Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
We aim to simulate random two-component vectors mimicking this structure. 
The margins are not normally distributed; particularly the Ozone level exhibits a strong positive skew.

```{r ch030-aq-joint-dist, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Bivariate scatterplot of Ozone vs. Temp with estimated marginal densities. We model the Ozone data as marginally log-normal and the Temperature data as normal.", cache=F}

@@

Next, we specify the marginal distributions and correlation coefficient (both type and magnitude).
Here the analyst is free to be creative.
For this example, we do not take up goodness-of-fit considerations to determine the marginal distributions.
But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.

*Specifying marginal distributions*.
Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed.
We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

@@

Next, we configure the input marginals for later input into `bigsimr::rvec`.
The marginal distributions are specified using `R`'s special `alist` function.
This allows one to enter the distributions without evaluating anything (yet).

@@

Notice that we use the *quantile* function for the marginals, as that is how the marginal distributions $F_i$ enter into the `bigsimr::rvec` algorithm.
This implementation strategy supports all `R` `base` probability distributions.
It allows flexible extensions using other `R` packages that adhere to conventions, such as `extraDistr`.
Further, by using `alist`, users can specify their own custom distributions (see below in [creating custom margins](#custom-margins)).

It is a bit inconvenient to repeatedly fill in the parameter values manually, so we provide a convenience function `mlist` which behaves similarly to `alist`, except that it evaluates the right side of argument values within the list.
This is intended to help with scaling up your code to high dimensions when many marginals to specify.

@@

*Specifying correlation*.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
These estimation methods are the standard approaches, not designed specifically designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

@@

*Checking the theoretical correlation bounds*.
As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
To ensure that the simulation is not configured to impossible settings, the `bigsimr::cor_bounds` function provides MC estimated theoretical lower and upper bounds (using the Generate, Sort, and Correlate algorithm of @DH2011).

@@

*Simulating random vectors*.
Finally, we arrive at the main function of `bigsimr`, `rvec`.
We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.

@@

*Creating a custom marginal distribution*.
Because `bigsimr` uses an `alist` to store the margins, any probability distribution with a well-defined inverse CDF can be employed, including custom marginal distributions not provided in base `R`. 
Therefore, to specify a marginal distribution absent from an existing package, the user must provide a closed-form expression to form the corresponding quantile function.

It is important to follow `R`'s naming convention for probability distributions.
Users should prefix distributions with `r` and `q` for *random* and *quantile* respectively.
Only the quantile function is necessary to simulate random vectors, but to compute the theoretical correlation bounds, one must to supply a univariate random number generator as well.

For an example, consider a custom *Pareto* distribution for use with `bigsimr`.
The Pareto CDF is

@@

Now with the `qpareto` and `rpareto` functions, we can use the distribution in `bigsimr` similar to the other built-in distributions.

@@

*Using `bigsimr` on a computing cluster via `rslurm`*.
Though `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server.
The R package `rslurm` makes it easy to run embarrassingly large parallel `rvec` calls.
This example assumes that `bigsimr` is installed on a system with a slurm scheduler.
A single run of `bigsimr:rvec` using `rslurm` can be coded as:

@@

The power of combining `bigsimr` and `rslurm` can be illustrated by simulating many correlation structures.
The `rslurm::slumr_map` syntax mirrors the `base::lapply` and `purrr::map` functions.

@@

On a cluster carrying 24 nodes with 48 threads, these 100 jobs complete in about a minute.
